Modal Abstractions for Virtualizing Memory Addresses

### 2023-03-23 programlogic.tex
+

    \[ \textsf{L4translate} \stackrel{\triangle}{=} \exists \entryf \ldotp \nfpointsto{\mask\vaddr\ft\crthree}{\mask\vaddr\tw\crthree}\entryf\qfracfotsss\naddr \]

@@

    \[ \textsf{L3translate} \stackrel{\triangle}{=} \exists \entrytr \ldotp \nfpointsto{\mask\vaddr\ft\entryf}{\mask\vaddr\tw\entryf}\entrytr\qfracfotss\naddr \]

@@

    \[ \textsf{L2translate} \stackrel{\triangle}{=} \exists \entrytw \ldotp \nfpointsto{\mask\vaddr\ft\entrytr}{\mask\vaddr\tw\entrytr}\entrytw\qfracfots\naddr \]

@@

    \[ \textsf{L1translate}  \stackrel{\triangle}{=} \exists \entryo \ldotp \nfpointsto{\mask\vaddr\ft\entrytw}{\mask\vaddr\tw\entrytw}\entryo\qfracfot\naddr \]

@@

    \[ \textsf{PageAccess} \stackrel{\triangle}{=} \exists \vpage \ldotp \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr \]

@@

\textsf{where   }\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\entryf,\entrytr,\entrytw,\entryo) \stackrel{\triangle}{=} \\
 \qquad\qquad \nfpointsto{\mask\vaddr\ft\crthree}{\mask\vaddr\tw\crthree}\entryf\qfracfotsss\naddr \ast \\ 
 \qquad\qquad  \nfpointsto{\mask\vaddr\ft\entryf}{\mask\vaddr\tw\entryf}\entrytr\qfracfotss\naddr  \ast  \\
  \qquad\qquad \nfpointsto{\mask\vaddr\ft\entrytr}{\mask\vaddr\tw\entrytr}\entrytw\qfracfots\naddr \ast \\
  \qquad\qquad \nfpointsto{\mask\vaddr\ft\entrytw}{\mask\vaddr\tw\entrytw}\entryo\qfracfot\naddr 
   \end{array}
\]
\caption{A Strong Virtual Points-to Relation}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:strongvirtualpointsto}
\end{figure*}
\end{remark}
%then we reach to a physical page addr (e.g. \textsf{pa}) 
%\[  \paddr\mapsto_{\textsf{a}}\{1\}\vsome \ast \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfrac\naddr \]

  \subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}
  We already build the intuition on what a virtual address mapping is composed of: 1. traversing the page tables to reach the page table entry 2. obtaining the physical address, and accessing the value mapped.

One might have already observed that the virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} is too strong to specify virtual memory operations updating a page table entry. Although there seems to fractional permissions for fragmental ownership of page-table entries, strong virtual points-to (in Figure \ref{strongvirtualpointsto}) relation doesn’t care about owning those fractional resources, it only cares that something ensures the correct page table walk exists. Any virtual memory operations such as mapping a new page which updates an page table entry would violate the soundness of other virtual-pointsto relations in the existence of page-sharing across the virtual addresses in the same address space.

One possible approach to is to just abstracting the \textit{valid} page-table walk existence from a each virtual address (e.g. \textsf{va}) to a physical page address, as \textit{per-element} ownership of a logical \textit{page-table-walk-summarization-map}
\[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  (\mathcal{W}_{64} \; \rightarrow_{\textrm{fin}}\;  (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
provided by a fragmentery ghost resrource ownership of $\mathcal{A}\textsf{bsPTableWalk}$.
\begin{remark}[Exchanging Page-Table-Walk-Summarization Mappings (Tokens) with Physical Page-Table-Walk Pointstos ]
$\mathcal{A}\textsf{PTableWalk}$ is a ghost map resource where fragmental-ownership of an individual key-value pair (per-element) can be handled as a resource. We use this mechanism to separate the virtual-to-physical translation from the physical memory involved: virtual points-to assertions can be defined in terms of fragmental ownership (ownership of an element) of these partial ghost map entry assertions , $\sumwalkabs\vaddr\qfrac\paddr$, plus sharing an invariant ($\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:peraspaceinvariant}) that the current installed page table respects all entries of the mapping, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$. From one another perspective, unmapping pages collects the fragmental map kvpairs, e.g. $\sumwalkabs\vaddr\qfrac\paddr$, from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical full ownership of the page tables lives, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$, in the central invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.% This still works with the modality, since that invariant is also semantically a predicate on a page table root.
\end{remark}
Then, our virtual memory addressing would be the fragmental ownership of the abstract page-table-walk mapping with the physical page address mapping as shown in Figure \ref{fig:virtualpointstosharing}. 
\begin{figure*}
\[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{v}}\{\textsf{q}\}\vpage \stackrel{\triangle}{=} 
  \exists \entryo \ldotp 
  \sumwalkabs\vaddr\qfrac\paddr \ast 
   \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr
\end{array}
\]
\caption{Virtual-Pointsto for Sharing Pages}
  \label{fig:virtualpointstosharing}
\end{figure*}

@@

@@

    \[ \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\vaddr,\entryf,\entrytr,\entrytw,\entryo)} \]

@@

  \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\crthree,\vaddr,\entryf,\entrytr,\entrytw,\entryo)}

@@

         [r](P)  \;\stackrel{\triangle}{=}  (\mathcal{W}_{64} -\textsf{b}> \textsf{uPredI} (\textsf{iResUR} \;\Sigma)) -\textsf{b}> \textsf{uPredI} \_ := \lambda \textsf{P},\; \textsf{P r}. 

>>>>

    \[ \begin{array}{l}
      \hbox{(\TirNameStyle{{L4translate}})} \quad
      \exists \entryf \ldotp \nfpointsto{\mask\vaddr\ft\crthree}{\mask\vaddr\tw\crthree}\entryf\qfracfotsss\naddr

@@

    \[\begin{array}{l}
    \hbox{(\TirNameStyle{{L3translate}})} \quad
    \exists \entrytr \ldotp \nfpointsto{\mask\vaddr\ft\entryf}{\mask\vaddr\tw\entryf}\entrytr\qfracfotss\naddr

@@

    \[ \begin{array}{l}
            \hbox{(\TirNameStyle{{L2translate}})} \quad 
       \exists \entrytw \ldotp \nfpointsto{\mask\vaddr\ft\entrytr}{\mask\vaddr\tw\entrytr}\entrytw\qfracfots\naddr

@@

    \[ \begin{array}{l}
      \hbox{(\TirNameStyle{{L1translate}})} \quad
      \exists \entryo \ldotp \nfpointsto{\mask\vaddr\ft\entrytw}{\mask\vaddr\tw\entrytw}\entryo\qfracfot\naddr

@@

    \[ \begin{array}{l}
      \hbox{(\TirNameStyle{{PageLebelAccess}})} \quad
      \exists \vpage \ldotp \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr

@@

\textsf{where   }\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr,\crthree,\entryf,\entrytr,\entrytw,\entryo) \stackrel{\triangle}{=} \\
 \qquad\qquad \nfpointsto{\mask\vaddr\ft\crthree}{\mask\vaddr\tw\crthree}\entryf\qfracfotsss\naddr \ast \\ 
 \qquad\qquad  \nfpointsto{\mask\vaddr\ft\entryf}{\mask\vaddr\tw\entryf}\entrytr\qfracfotss\naddr  \ast  \\
  \qquad\qquad \nfpointsto{\mask\vaddr\ft\entrytr}{\mask\vaddr\tw\entrytr}\entrytw\qfracfots\naddr \ast \\
  \qquad\qquad \nfpointsto{\mask\vaddr\ft\entrytw}{\mask\vaddr\tw\entrytw}\entryo\qfracfot\naddr 
   \end{array}
\]
\caption{A Strong Virtual Points-to Relation}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:strongvirtualpointsto}
\end{figure*}
\end{remark}
%then we reach to a physical page addr (e.g. \textsf{pa}) 
%\[  \paddr\mapsto_{\textsf{a}}\{1\}\vsome \ast \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfrac\naddr \]

  \subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}  
  One might have already observed that the virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} is too strong to specify virtual memory operations updating a page table entry. To speak more concretely, the fractional permissions on the page-table-walk assertion (\textsf{L}$_4$\_\textsf{L}$_{1}$ in Figure \ref{fig:strongvirtualpointsto}) \textit{only} assure the existence of the certain number of mappings -- 512 at each level -- abstracting virtual address translations, and updating any of these mappings would make the other mappings, address translations sharing the page table mappings. Therefore, we need to have logical abstractions ensuring sound updates to the page-table entries in the existence of sharing.

 One possible approach is to abstract away the physical page-table-walk pointsto (\textsf{L}$_4$\_\textsf{L}$_{1}$ in Figure \ref{fig:strongvirtualpointsto}) from the virtual pointsto relation via replacing it with read-only abstract map entries -- a ghost (abstract) map entry per \textsf{L}$_4$\_\textsf{L}$_{1}$ mappings -- which can be used as a resource ($ \sumwalkabs\vaddr\qfrac\paddr$) whose ownership ensures that the virtual address ($\vaddr$) is mapped to a page (referenced by $\paddr$) ,and as a token to access to the physical page-table-walk pointsto (\textsf{L}$_4$\_\textsf{L}$_{1}$) stored in an address-space invariant. We use \textsf{Iris}'s authorative ghost-map logical machinery
\[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  (\mathcal{W}_{64} \; \rightarrow_{\textrm{fin}}\;  (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
to realize these reasoning principles ($\mathcal{A}\textsf{bsPTableWalk}$).
\begin{remark}[Exchanging Page-Table-Walk-Summarization Mappings (Tokens) with Physical Page-Table-Walk Pointstos ]
$\mathcal{A}\textsf{PTableWalk}$'s fragmental-ownership represents an individual key-value pair (per-element) which can be handled as a resource. Critically, physical full ownership of the page tables lives, $\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$, in the central invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables. We use this mechanism to separate the virtual-to-physical translation from the physical memory involved: virtual points-to assertions can be defined in terms of fragmental ownership (ownership of an element) of these partial ghost map entry assertions , $\sumwalkabs\vaddr\qfrac\paddr$, and physical page mapping ($nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr$). Physical page-table-walk pointsto, unlike the case in the strong virtual-pointsto relation in Figure \ref{fig:strongvirtualpointsto}, is moved into an invariant ($\mathcal{I}$\textsf{ASpace} in Figure \ref{fig:peraspaceinvariant}), and resides with authorative ownership of $\mathcal{A}\textsf{PTableWalk}$ ($\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}$) which respects all entries of the abstract virtual-address translation mappings ($\sumwalkabs\vaddr\qfrac\paddr$). To build more intuition, we can see the use-case \textit{unmapping-pages} as collecting the fragmental map pairs, e.g. $\sumwalkabs\vaddr\qfrac\paddr$, from collecting the assertions, removing them from the ghost map, and then updating the page tables. % This still works with the modality, since that invariant is also semantically a predicate on a page table root.
\end{remark}
Then, our virtual memory addressing would be the fragmental ownership of the abstract page-table-walk mapping with the physical page address mapping as shown in Figure \ref{fig:virtualpointstosharing}. 
\begin{figure*}
\[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{v}}\{\textsf{q}\}\vpage \stackrel{\triangle}{=} 
  \exists \entryo \ldotp 
  \sumwalkabs\vaddr\qfrac\paddr \ast 
   \nfpointsto{\mask\vaddr\ft\entryo}{\mask\vaddr\tw\entryo}\vpage\qfracone\naddr
\end{array}
\]
\caption{Virtual-Pointsto for Sharing Pages}
  \label{fig:virtualpointstosharing}
\end{figure*}

@@

   \includegraphics[width=0.75\columnwidth]{logical_addr_space.png}
  \caption{Logical Resources of Address-Space Abstraction}
  \label{fig:logicaladdrspace}

@@

    \[ \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr,\crthree,\entryf,\entrytr,\entrytw,\entryo)} \]

@@

  \bigast{(\vaddr, \entryo)\in \ptablestore}{\exists\;(\entryf\;,\entrytr\;,\entrytw)\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr,\crthree,\entryf,\entrytr,\entrytw,\entryo)}

@@

         [r](P)  \;\stackrel{\triangle}{=}  (\mathcal{W}_{64} \rightarrow_{\textsf{b}} \textsf{uPredI} (\textsf{iResUR} \;\Sigma)) \rightarrow_{\textsf{b}} \textsf{uPredI } \_ := \lambda \textsf{P},\; \textsf{P r}. 
